Draft: Not So Bad---The minimal effect of election fraud on protest intensity in non-democracies

### 2022-08-26 Output/manipulation_protest_draft_v2.Rmd
First draft appendix

---
title: "Draft: It's Not So Bad---The minimal effect of election fraud on protest intensity in non-democracies"
author: "Cole J. Harvey (Oklahoma State University)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
    bookdown::pdf_document2: default
linkcolor: blue
bibliography: C:/Users/colej/Documents/Research projects/bib_ch.bib
#biblio-style: "apalike"
fontsize: 12pt
link-citations: yes
toc: false
csl: C:/Users/colej/Documents/Research projects/ji_manipulation_protest/american-journal-of-political-science.csl
header-includes: \usepackage{booktabs}
abstract: "The risk of mass protest figures prominently in many models of election manipulation, as a deterrent to tampering with elections. Yet election manipulation is common in non-democracies, while protest is rare, and successful protest rarer still. Using count data on protest events from 647 elections from 1990 to 2012, this paper argues that the risk to incumbents has been overstated in prior work. Specifically, by treating protest as a binary event, previous research has overestimated the strength of protest movements relative to the incumbent's institutional resources. The results show that the level of fraud has no bearing on the number of protests after an election, though protest initiation is more likely in high-fraud elections with low margins of victory. This pessimistic finding implies that, rather than serving as a constraint, the risk of protest creates a perverse incentive to manipulate as much as possible in order to achieve a large official vote-share."
---

@@

Using cross-national data on 647 elections from 1990 to 2012, adapted from the recently-released Electoral Contention and Violence (ECAV) dataset [@Daxecker2019], this paper finds that more fraudulent elections are associated with a higher risk that any post-election protest event occurs. This is in keeping with the formal-modeling literature, which predicts that overreach in the form of election fraud indicates government weakness, making opposition groups confident of winning a conflict in the streets. However, there is no statistically significant association between the level of election fraud and the number of protest events. This suggests a more complex relationship between election fraud and protest.

Since fraud is not associated with the overall number of events in a protest wave, the risk to governments has likely been overemphasized. That is, even if protests do break out in response to large-scale fraud, the government has many tools it may deploy to curtail them. Once protests begin, the noisy signal of regime strength provided by the election is replaced by a very clear signal: the force of the regime's response to the emerging protest movement in contest with the mobilizational capacity of the opposition.  To the extent that larger protest waves are more damaging to the government than small ones, protest risk does not appear to be a substantial constraint on the severity of election fraud. Troublingly, these findings indicate that ruling parties should follow an 'all-in' strategy on election manipulation. More manipulation is not associated with more protest mobilization, but a large margin of victory is associated with less risk of protest initiation. These results help explain both the relative paucity of electoral protests, as well as the rarity of their success in overturning incumbent regimes.

@@

Despite this consensus in the formal literature, there is relatively little empirical evidence to support the claim that manipulation per se indicates weakness, and thus increases protest risk. On the contrary, Simpser [-@Simpser2011] shows support for the alternative (though still informational) view that excessive manipulation can induce cooperation from opposition groups, elites, and other actors by revealing information about regime strength. Similarly, the type of manipulation employed can convey information about the ruling party's capacity to mobilize [@Harvey2020]. In this way, large-scale manipulation can contribute to a ruling party's ability to produce a large, stabilizing majority vote-share [@magaloni2006]. Similarly, Luo and Rozenas [-@Luo2020] posit that ex post fraud is more likely to result in protest than ex ante manipulation, though this can be contrasted with the finding that parties are more likely to protest elections after ex ante changes to the rules [@chernykh2014]. In sum, it is likely that manipulation *may* signal regime weakness, but this is more likely when margins of victory are narrow and when falsification is a major tool.

@@

For example, ballot-stuffing and voter coercion have been shown to generate stronger feelings of anger than pre-election manipulation of ballot access [@Szakonyi2021], and access to information about manipulation appears to reduce support for the ruling party [@Reuter2021], increase support for protest and reduce the perceived legitimacy of the regime [@Williamson2021]. This argument has found support in studies of the relationship between economic grievances, manipulated elections, and protest [@Rod2019;@Brancati2016]. A study of individuals in Nigeria found that individuals' subjective perceptions of fraud makes them more likely to participate in protest, not more objective measures [@Daxecker2019b], which is suggests that emotions, moral condemnation, or grievances take precedence over strategic calculations for individuals. Subjective belief in fraud was also shown to be correlated with approval of mass protests after Russia's 2011 election [@Chaisty2013]. Furthermore, simply being on the losing side in an election is associated with reduced belief that the election was fair [@Cantu2015], and with increased support for electoral protest [@Anderson2006]. [Note: may add procedural justice cites to this graf.]

These two explanations are, of course, interconnected. As Schedler writes, "Grievances provide moral justifications for collective protest: we need to do something. Opportunities provide strategic justifications: we can achieve something" [@Schedler2013,p.304]. Several empirical studies have, accordingly, found links between election manipulation and binary measures of protest [@Brancati2016;@Hyde2014a;@Harvey2020;@Rod2019].

The literature thus expects that manipulated elections increase the risk of protest by 1) revealing regime weakness, especially when the margin of victory is low, and 2) exacerbating existing grievances and upsetting citizens' sense of procedural fairness. However, as discussed below, electoral protest is relatively rare, and even in cases where it does occur, most efforts to overturn fraudulent election results through protest have failed [@kalandadze2009electoral]. The apparent failure of protest risk to deter election fraud underlines our limited understanding of non-democratic ruling parties' decision-making. Does the extent of fraud, or the severity of underlying social grievances, make more numerous (and thus more costly) protest more likely?

@@

>H2: Post-election protest will be more likely to occur when fraud is severe and ruling-party margin of victory is small, and less likely when fraud is low or the margin of victory large.

@@

However, these predictions may not apply to the size or intensity of a protest wave. Once protests begin, the extent of electoral fraud is unlikely to affect their scope. The initial emergence of protest triggers a response by the government and ruling party that can take the form of repression, concessions, or toleration [@Tilly1977;@Lipsky1968;@Hummel2019;@Rasler1996;@Bishara2015;@Frantz2014]. (**See others**.) Given that post-electoral protests are a direct challenge to the regime, toleration is unlikely in these cases. That election dates are known in advance creates opportunities for opposition mobilization; anticipating this, regimes can more easily engage in pre-emptive repression [@Truex2019;@Sullivan2016;@Bhasin2013]. Because the emergence of protest indicates that the regime's pre-emptive efforts have failed, it is not clear a priori how the regime should handle protesters it failed to deter [@Ritter2016]. In many cases, the government doubles down on repression, calling upon the security services to disperse protests, arrest leaders, and perhaps employ harsher methods including disappearance, torture, and execution [@Escriba_Folch2013;@Davenport2007]. Though such crackdowns risk provoking a backlash, they are aimed at demobilizing the opposition by increasing the costs of participation in overt dissent [@DeVogel2022]. In other cases, the government may attempt to accommodate the opposition, by making concessions on policy or institutions [@Moore2000;@Hafner-Burton2016;@Piven1979]. Alternatively, concessions may be aimed at the more moderate public, in order to keep them on the sidelines of the conflict [@Hummel2019].

What results is a test of strength between the anti-incumbent protest coalition and the regime with its supporters (Tilly 2006). In essence, as Beaulieu [@Beaulieu2014a] notes, the onset of electoral protest marks a case of bargaining failure. Just as in the international context, the ensuing conflict acts to reveal information on the strength and resolve of the two sides.  In such a contest, the incumbent enjoys multiple advantages, though it may lack the support of a majority of the population. It controls the resources of the state, and can call upon the support of those elites who benefit from the status quo. High-capacity states, as a result, are more likely to be able to engage in successful repression (Tarrow 2011). Though a biased election may temporarily allow diverse opposition groups to set aide their differences [@trejo_ballot_2014], this internal diversity can enable the incumbent to divide the opposition with selective repression or concessions. For example, survey research from Russia's 'For Fair Elections' movement found that many protesters were motivated as much by... [Find citation]. While the regime may have its own splits---between reformists and hardliners, for example---the emergence of protest can act to solidify support for the ruling party among its supporters.

@@

These hypotheses are tested using data from three datasets: NELDA [@hyde2012], V-Dem version 11.1 [@vdem11.1], and the Electoral Contention and Violence (ECAV) dataset [@Daxecker2019]. The unit of observation for this study is an election period as recorded in NELDA, which counts all national elections between 1945 and 2015. This framework prevents selecting on the dependent variable, by capturing both protested and non-protested elections. NELDA also includes a variable indicating elections in which incumbents lost; as this study is focused on protest against ostensibly victorious incumbents, I use this variable to filter out elections which the opposition won outright. I further limit the sample to regimes coded as electoral autocracies or electoral democracies in V-Dem. This excludes closed autocracies (where no national elections are held) and liberal democracies (where electoral integrity and rule of law are both high). Finally, the date range is limited by the availability of data on protest from ECAV, as discussed below, which runs from 1990 to 2012. The resulting dataset includes 647 election-period observations.

@@

Each model includes the explanatory variables proposed above. First, three proxy measures of economic grievance include an indicator of *public sector corruption* and the *lagged GDP growth rate*, both taken from V-Dem, and a binary measure of *economic crisis* from the NELDA dataset. Higher levels of corruption, negative GDP growth rates, and a positive value for *economic crisis* should each be expected to increase citizens' sense of grievance. To capture the *severity of election fraud*, I use the V-Dem measure of intentional election irregularities, a category which includes "double IDs, intentional lack of voting materials, ballot-stuffing, misreporting of votes, and false collation of votes." I operationalize electoral manipulation using this variable, rather than aggregate measures of election integrity, because fraud of this kind is thought to be most likely to trigger protest [@Linebarger2020]---it is often incumbents' tool of last resort [@Sjoberg2016], it can fail to signal strength [@Harvey2020], and it can be more damaging to legitimacy than pre-election manipulation [@birch2011;@Szakonyi2021]. Finally, the winning party's *vote-share* is also taken from the V-Dem dataset.

@@

Turning to models of protest quantity presented in Table 3, there is no statistically significant relationship between fraud and protest in any of the models. Figure 4, using results from Model 9, presents the marginal effects for fraud conditional on the incumbent's vote-share. As the figure shows, there is no statistically significant difference in the expected number of protest following a relatively clean election compared to a relatively fraudulent one, for any margin of incumbent victory. Taking the results from Tables 2 and 3 together, there is support for Hypothesis 1 (an unconditional increase in the risk of protest as fraud increases), as well as strong support for Hypothesis 2 (an increase in protest risk in fraudulent, low-margin elections) and Hypothesis 4 (no relationship between fraud and protest quantity). Hypothesis 3, on the connection between fraud, grievance, and protest, is not supported.

@@

Altogether, the findings support an understanding of fraud as a signal of incumbent strength, but a noisy and unreliable one. When an incumbent party wins by a narrow margin in a fraudulent election, the likelihood of protest initiation is substantially higher than when it wins more cleanly: roughly a 75\% chance in the former case versus 15\% in the latter as shown in Figure 2. However, since even strong incumbents commit fraud, the decision to take to the streets is often a losing gamble---anticipating a weak incumbent, they instead find one that is capable of repressing, coopting, or conciliating an emerging protest movement. Indeed, while there are 187 cases of protest out of the 647 elections in this dataset, only 15 elections resulted in either the incumbent being replaced due to protest or the election being re-run.\footnote{This information is determined using the NELDA variables NELDA37 and NELDA41. The cases of successful protest are: Ukraine 2004, Azerbaijan 2005, Georgia 2003, Kazakhstan 1994, Kyrgyzstan 2005, Bulgaria 1990, Malaysia 2004, Bangladesh 1996, Mali 1997, Peru 2000, Burkina Faso 1997, and Cote d'Ivoire 2000.} This effect can be seen in the results from the quantity models.

The above results have several implications for understanding election manipulation and regime resilience.  Most importantly, by treating protest primarily as a binary event rather, the existing literature on electoral protest has overemphasized the risk to incumbents of engaging in electoral manipulation. This results presented here show that most electoral protests quickly dwindle, and that fraud has no bearing on how numerous protest events are. Moreover, contrary to some prior research, the risk of protest does not appear to be heightened by poor economic conditions [@Brancati2014a] or intensive corruption in the public sector. As a result, circumstances that are to some extent outside of leaders' control, like a worsening macroeconomic climate, do not appear to make engaging in fraud any riskier. Instead, the conditions that are correlated with more damaging protest are generally institutional factors, which are either more amenable to control or can be anticipated by non-democratic incumbents.  Legislative constraints, an important factor for the duration and number of protests, can be adjusted through institutional reforms and cooptation of the opposition, over time. Finally, incumbent term limits can be managed through constitutional reforms, where possible, or through the appointment of (presumably) loyal successors. In other words, election fraud only has a minor cost in terms of protest risk, a risk which can be further blunted to the extent that elections can often be held on incumbents' terms. These are pessimistic findings for proponents of democracy.

@@

None of this is to say that election fraud cannot, in theory, pose a cost to incumbents in the form of protest. This study is not a causal one; the decision by the incumbent to engage in fraud and the decision by opposition groups to protest are endogenous to one another. Instead, it shows that in practice, manipulated elections are not associated with more numerous protests than clean ones. In typical cases, as a result, protest risk is likely not the constraining factor on the level of election fraud, contrary to the assumption in much of the formal theoretical literature. Rather, the constraints on manipulation are more likely to be driven by factors like resource availability [@greene_why_2007], principal-agent problems [@rundlett2016deliver], and legal risks to manipulators [@Harvey2022].

>>>>

---
title: "Draft: Not So Bad---The minimal effect of election fraud on protest intensity in non-democracies"
author: "Cole J. Harvey (Oklahoma State University)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
    bookdown::pdf_document2: default
linkcolor: blue
bibliography: C:/Users/colej/Documents/Research projects/bib_ch.bib
#biblio-style: "apalike"
fontsize: 12pt
link-citations: yes
toc: false
csl: C:/Users/colej/Documents/Research projects/ji_manipulation_protest/american-journal-of-political-science.csl
header-includes: \usepackage{booktabs}
abstract: "The risk of mass protest figures prominently in many models of election manipulation, as a deterrent to tampering with elections. Yet election manipulation is common in non-democracies, while protest is rare, and successful protest rarer still. Using count data on protest events from 647 elections from 1990 to 2012, this paper argues that the risk to incumbents has been overstated in prior work. Specifically, by treating protest as a binary event, previous research has overestimated the strength of protest movements relative to the incumbent's institutional resources. The results show that the level of fraud has no bearing on the number of protests after an election, though protest initiation is more likely in high-fraud elections with low margins of victory. This pessimistic finding implies that, rather than serving as a constraint, the risk of protest creates a perverse incentive to manipulate as much as possible in order to achieve a large official vote-share."
---

@@

Using cross-national data on 647 elections from 1990 to 2012 where incumbents claimed victory, adapted from the recently-released Electoral Contention and Violence (ECAV) dataset [@Daxecker2019], this paper finds support for the hypothesis that more fraudulent elections are associated with a higher risk that any post-election protest event occurs. This is in keeping with the formal-modeling literature, which predicts that overreach in the form of election fraud indicates government weakness, making opposition groups confident of winning a conflict in the streets. However, there is no statistically significant association between the level of election fraud and the number of protest events. This suggests a more complex relationship between election fraud and protest, undermining the argument that protest risk is a major deterrent to electoral malfeasance.

Since fraud is not associated with the overall number of events in a protest wave, the risk to governments has likely been overemphasized. That is, even if protests do break out in response to large-scale fraud, the government has many tools it may deploy to curtail them. Once protests begin, the noisy signal of regime strength provided by the election is replaced by a very clear signal: the force of the regime's response to the emerging protest movement in contest with the mobilizational capacity of the opposition.  To the extent that larger protest waves are more damaging to the government than small ones, protest risk does not appear to be a substantial constraint on the severity of election fraud. Troublingly, these findings indicate that ruling parties should follow an 'all-in' strategy on election manipulation. More manipulation is not associated with a larger number of protests, but a large margin of victory is associated with a lower risk of protest initiation. These results help explain both the relative paucity of electoral protests, the rarity of their success in overturning incumbent regimes, and governments' attempts to claim implausibly high margins of victory.

@@

Despite this consensus in the formal literature, there is relatively little empirical evidence to support the claim that manipulation per se indicates weakness, and thus increases protest risk. On the contrary, Simpser [-@Simpser2011] shows support for the alternative view that excessive manipulation can induce cooperation from opposition groups, elites, and other actors by revealing information about regime strength. Similarly, the type of manipulation employed can convey information about the ruling party's capacity to mobilize, with falsification providing the least credible information [@Harvey2020]. In this way, large-scale manipulation can contribute to a ruling party's ability to produce a large, stabilizing majority vote-share [@magaloni2006]. Similarly, Luo and Rozenas [-@Luo2020] posit that ex post fraud is more likely to result in protest than ex ante manipulation, though this can be contrasted with the finding that parties are more likely to protest elections after ex ante changes to the rules [@chernykh2014]. In sum, it is likely that manipulation *may* signal regime weakness, but this is more likely when margins of victory are narrow and when falsification is a major tool.

@@

For example, ballot-stuffing and voter coercion have been shown to generate stronger feelings of anger than pre-election manipulation of ballot access [@Szakonyi2021], and access to information about manipulation appears to reduce support for the ruling party [@Reuter2021], increase support for protest and reduce the perceived legitimacy of the regime [@Williamson2021]. This argument has found support in studies of the relationship between economic grievances, manipulated elections, and protest [@Rod2019;@Brancati2016]. A study of individuals in Nigeria found that individuals' subjective perceptions of fraud makes them more likely to participate in protest, not more objective measures [@Daxecker2019b], which suggests that emotions, moral condemnation, or grievances take precedence over strategic calculations for individuals. Subjective belief in fraud was also shown to be correlated with approval of mass protests after Russia's 2011 election [@Chaisty2013]. Furthermore, simply being on the losing side in an election is associated with reduced belief that the election was fair [@Cantu2015], and with increased support for electoral protest [@Anderson2006]. [Note: may add procedural justice cites to this graf.]

The strategic and social-psychological explanations are, of course, interconnected. As Schedler writes, "Grievances provide moral justifications for collective protest: we need to do something. Opportunities provide strategic justifications: we can achieve something" [@Schedler2013,p.304]. Several empirical studies have, accordingly, found links between election manipulation and binary measures of protest [@Brancati2016;@Hyde2014a;@Harvey2020;@Rod2019].

The literature thus expects that manipulated elections increase the risk of protest by 1) revealing regime weakness, especially when the margin of victory is low, and 2) exacerbating existing grievances and upsetting citizens' sense of procedural fairness. In turn, governments should be reluctant to manipulate. However, as discussed below, electoral protest is relatively rare, and even in cases where it does occur, most efforts to overturn fraudulent election results through protest have failed [@kalandadze2009electoral]. The apparent failure of protest risk to deter election fraud underlines our limited understanding of non-democratic ruling parties' decision-making. Does the extent of fraud, in fact, make more numerous (and thus more costly) protest more likely?

@@

>H2: More sever election fraud is associated with a higher likelihood of protest when winning-party vote-share is low.

@@

However, these predictions may not apply to the size or intensity of a protest wave. Once protests begin, the extent of electoral fraud is unlikely to affect their scope. The initial emergence of protest triggers a response by the government and ruling party that can take the form of repression, concessions, or toleration [@tillyMobilizationRevolution1978;@lipskyProtest1968;@Hummel2019;@raslerConcessionsRepression1996;@Bishara2015;@Frantz2014]. Given that post-electoral protests are a direct challenge to the regime, toleration is unlikely in these cases. That election dates are known in advance creates opportunities for opposition mobilization; anticipating this, regimes can more easily engage in pre-emptive repression [@Truex2019;@Sullivan2016;@Bhasin2013]. Because the emergence of protest indicates that the regime's pre-emptive efforts have failed, it is not clear a priori how the regime should handle protesters it failed to deter [@Ritter2016]. In many cases, the government doubles down on repression, calling upon the security services to disperse protests, arrest leaders, and perhaps employ harsher methods including disappearance, torture, and execution [@Escriba_Folch2013;@Davenport2007]. Though such crackdowns risk provoking a backlash, they are aimed at demobilizing the opposition by increasing the costs of participation in overt dissent [@DeVogel2022]. In other cases, the government may attempt to accommodate the opposition, by making concessions on policy or institutions [@Moore2000;@Hafner-Burton2016;@Piven1979]. Alternatively, concessions may be aimed at the more moderate public, in order to keep them on the sidelines of the conflict [@Hummel2019].

What results is a test of strength between the anti-incumbent protest coalition and the regime with its supporters [@mcadamDynamicsContention2001;@tillyRegimesRepertoires2010]. In essence, as Beaulieu [@Beaulieu2014a] notes, the onset of electoral protest marks a case of bargaining failure. Just as in the international context, the ensuing conflict acts to reveal information on the strength and resolve of the two sides.  In such a contest, the incumbent enjoys multiple advantages, though it may lack the support of a majority of the population. It controls the resources of the state, and can call upon the support of those elites who benefit from the status quo. High-capacity states, as a result, are more likely to be able to engage in successful repression [@tarrowPowerMovementSocial2011]. Though a biased election may temporarily allow diverse opposition groups to set aide their differences [@trejo_ballot_2014], this internal diversity can enable the incumbent to divide the opposition with selective repression or concessions. While the regime may have its own splits---between reformists and hardliners, for example---the emergence of protest can act to solidify support for the ruling party among its supporters.

@@

These hypotheses are tested using data from three datasets: NELDA [@hyde2012], V-Dem version 11.1 [@vdem11.1], and the Electoral Contention and Violence (ECAV) dataset [@Daxecker2019]. The unit of observation for this study is an election period as recorded in NELDA. This framework prevents selecting on the dependent variable, by capturing both protested and non-protested elections. NELDA also includes a variable indicating elections in which incumbents lost; as this study is focused on protest against ostensibly victorious incumbents, I use this variable to filter out elections which the opposition won outright. I further limit the sample to regimes coded as electoral autocracies or electoral democracies in V-Dem. This excludes closed autocracies (where no national elections are held) and liberal democracies (where electoral integrity and rule of law are both high). Finally, the date range is limited by the availability of data on protest from ECAV, as discussed below, which runs from 1990 to 2012. The resulting dataset includes 647 election-period observations across 109 countries.

@@

Each model includes the explanatory variables proposed above. First, three proxy measures of economic grievance include an indicator of *public sector corruption* and the *lagged GDP growth rate*, both taken from V-Dem, and a binary measure of *economic crisis* from the NELDA dataset. Higher levels of corruption, negative GDP growth rates, and a positive value for *economic crisis* should each be expected to increase citizens' sense of grievance. To capture the *severity of election fraud*, I use the V-Dem measure of intentional election irregularities, a category which includes "double IDs, intentional lack of voting materials, ballot-stuffing, misreporting of votes, and false collation of votes." I operationalize electoral manipulation using this variable, rather than aggregate measures of election integrity, because fraud of this kind is thought to be most likely to trigger protest [@Linebarger2020]---it is often incumbents' tool of last resort [@Sjoberg2016], it can fail to signal strength [@Harvey2020], and it can be more damaging to legitimacy than pre-election manipulation [@birch2011;@Szakonyi2021]. This measure of election integrity should thus pose a strong test of the protest-oriented model. Finally, the winning party's *vote-share* is also taken from the V-Dem dataset.

@@

Turning to models of protest quantity presented in Table 3, there is no statistically significant relationship between fraud and protest in any of the models. Figure 4, using results from Model 9, presents the marginal effects for fraud conditional on the incumbent's vote-share. As the figure shows, there is no statistically significant difference in the expected number of protest following a relatively clean election compared to a relatively fraudulent one, for any margin of incumbent victory. Taking the results from Tables 2 and 3 together, there is support for Hypothesis 1 (an unconditional increase in the risk of any protest as fraud increases), as well as strong support for Hypothesis 2 (an increase in protest risk in fraudulent, low-margin elections) and Hypothesis 4 (no relationship between fraud and protest quantity). Hypothesis 3, on the connection between fraud, grievance, and protest, is not supported.

@@

Altogether, the findings challenge the idea that the expectation of protest will deter election-manipulation by ruling parties. To the extent that fraud is a signal of incumbent strength, it is a noisy and unreliable one. When an incumbent party wins by a narrow margin in a fraudulent election, the likelihood of protest initiation is substantially higher than when it wins more cleanly: roughly a 75\% chance in the former case versus 15\% in the latter as shown in Figure 2. However, since even strong incumbents commit fraud, the decision to take to the streets is often a losing gamble---anticipating a weak incumbent, they instead find one that is capable of repressing, coopting, or conciliating an emerging protest movement. Indeed, while there are 187 cases of protest out of the 647 elections in this dataset, only 15 elections resulted in either the incumbent being replaced due to protest or the election being re-run.\footnote{This information is determined using the NELDA variables NELDA37 and NELDA41. The cases of successful protest are: Ukraine 2004, Azerbaijan 2005, Georgia 2003, Kazakhstan 1994, Kyrgyzstan 2005, Bulgaria 1990, Malaysia 2004, Bangladesh 1996, Mali 1997, Peru 2000, Burkina Faso 1997, and Cote d'Ivoire 2000.} This effect can be seen in the results from the quantity models.

The above results have several implications for understanding election manipulation and regime resilience.  Most importantly, by treating protest primarily as a binary event rather, the existing literature on electoral protest has overemphasized the risk to incumbents of engaging in electoral manipulation. This results presented here show that most electoral protests quickly dwindle, and that fraud has no bearing on how numerous protest events are. Moreover, contrary to some prior research, the risk of protest does not appear to be heightened by poor economic conditions [@Brancati2014a] or intensive corruption in the public sector. As a result, circumstances that are to some extent outside of leaders' control, like a worsening macroeconomic climate, do not appear to make engaging in fraud any riskier. Instead, the conditions that are correlated with more damaging protest are generally institutional factors, which are either more amenable to control or can be anticipated by non-democratic incumbents.  Legislative constraints, an important factor for the duration and number of protests, can be adjusted through institutional reforms and cooptation of the opposition, over time. Incumbent term limits can be managed through constitutional reforms, where possible, or through the appointment of (presumably) loyal successors. In other words, election fraud only has a minor cost in terms of protest risk, a risk which can be further blunted to the extent that elections can often be held on incumbents' terms. These are pessimistic findings for proponents of democracy.

@@

One objection to this framing may be that, if the level of manipulation and the risk of protest are endogenous to one another, perhaps the null result found for the number of protests is to be expected. That is, if incumbents choose a level of manipulation calculated to avoid large-scale protest given other contextual factors, no correlation between fraud and protest quantity would be found. This paper does not dispute that decisions about manipulation and protest may be endogenous, but it does argue that the resulting equilibrium favors governments. In the dataset used here, only ten elections in four countries are coded as having an *intentional irregularities* score of less than -2 (indicating extremely fraud-free elections). By contrast, 103 elections in 27 countries score higher than 2 (indicating highly fraudulent elections). The frequency of fraud, the rarity of protest, and the finding that fraud severity is not correlated with the scope of protest events all point to the same conclusion: protest risk is not a substantial constraint on election fraud in non-democracies. Rather, the constraints on manipulation are more likely to be driven by factors like resource availability [@greene_why_2007], principal-agent problems [@rundlett2016deliver], and legal risks to manipulators [@Harvey2022].
